Omics-based Single Step Trait Prediction

### 2017-03-30 paper/manuscript.tex
Reorder paragraphs in discussion

\section*{Discussion}
\Genetics2level{Advantages of marker effect models}
Single step approaches in hybrid populations have so far been limited to pig
breeding \cite{Xiang2015,Xiang2016,Tusell2016}.
All of these studies used breeding value-based single step BLUP methods
(SS-BLUP) conceived by \citeNP{Legarra2009} and \citeNP{Christensen2010} with
extensions to hybrid populations \cite{Christensen2014,Christensen2015}.
An attactive property of SS-BLUP models is their computational advantage over 
marker-effect single step Bayesian regression models (SSBR) when the number of 
genotypes is smaller than the number of features of the incomplete predictor.
Nevertheless, SS-BLUP suffers from a variety of issues, which we will present
briefly:
First, the combination of the numerator with the genomic relationship matrix
requires commensurability of the two matrices, which necessitates some form of
scaling and weighting of $\mathbf{A}$ and $\mathbf{G}$
\cite{Christensen2012,Christensen2012a}.
Second, the fact that mostly recent individuals have genotypes whereas the 
majority of old individuals has only pedigree records introduces a bias that 
has to be accounted for \cite{Vitezica2011,Legarra2015,Garcia-Baccino2017}.
Third, the addition of new genotypes requires updates of $\mathbf{G}$ that
demand additional measures such as computing the inverse of $\mathbf{G}$ via
recursion
\cite{Misztal2014,Misztal2016,Misztal2016a,Fragomeni2015,Masuda2016,Pocrnic2016}.
Advantages of the SSBR algorithm, conceived by \citeNP{Fernando2014}, include
that i) it does not require commensurability of $\mathbf{A}$ and $\mathbf{G}$, 
ii) it models the bias incurred by the incomplete predictor explicitly as 
$\mathbf{\epsilon}$ and iii) its speed depends largely on the number of 
features.
To our best knowledge, this study presents the first implementation of the SSBR 
algorithm in a hybrid data set.

@@

\Genetics2level{Superiority of single step models}
\subsection{Inbred lines}
Predictive abilities for the maize diversity panel are not in complete
agreement with those reported by \citeNP{Guo2016}.
Reasons for this may be threefold: 
1) In our study, we reduced the dataset to the 211 tropical/subtropical lines 
and excluded all other genotypes whereas \citeNP{Guo2016} used genotypes from 
four subpopulations and only a subset of 157 genotypes from the set of 
tropical/subtropical lines while correcting for population structure by
including a fixed effect in the model for predicting agronomic performance.
2) We applied quality checks for the predictor data after generating the subset
of 211 inbred lines, whereas \citeNP{Guo2016} applied the quality checks to 368
genotypes.
3) While \citeNP{Guo2016} used five-fold cross-validation with 500 repetitions,
we employed a leave-one-out cross-validation scheme (LOOCV).
When using a single step approach based on complete genomic and incomplete
transcriptomic information, gains in predictive ability for the inbred lines
were at best just slightly higher than when using complete genomic information
alone.
One of the reasons might be that the subset of genotypes that was covered by
transcriptomic data was not a good representation of the full genetic space of
all available genotypes.
A second possibility is that the number of additional phenotypes in the full
set compared to the core set ($\Delta(n) = 62$) was not large enough to provide
considerably more useful information.
The latter was the case for \citeNP{Ashraf2016}, who imputed genotypes for about 
10,000 wheat lines using pedigree information and observed greater accuracy of
the single step method over genomic BLUP for all four evaluated traits.

\subsection{Maize hybrids}
Hybrid breeding --- with noteworthy commercial applications in pig, chicken and 
maize breeding --- is a particularly challenging application field for
prediction tools.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology \cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.
% Are there applications in chicken breeding? (Hints in Christensen et al.
% 2014, GSE).
For the hybrid data, the predictive abilities of genomic information
for the whole set of available parent lines rose considerably for several
agronomic traits.
Hence, the inclusion of trancsriptomic data could not further improve upon
this result.
Predictive abilities obtained when using pedigree data could be 
improved considerably when, depending on the agronomic trait, combined with 
either transcriptomic or genomic information.
Evaluations of core sets for the parental inbred lines of the maize hybrids
used throughout this study were outside the scope of this paper, but
results on the inbred lines have shown that homogeneous coverage of the
genetic space by the incomplete predictor can considerably leverage the
predictive ability in single step prediction compared to using only the
complete predictor.
In hybrid breeding, the probability of successfully selecting the best
observed genotypes based on the best predicted candidates is a strongly 
convex function of the predictive ability.
Hence, even the minor gains when using transcriptomic and genomic data in a
single step might justify additional investments in RNA-seq at least of a
subset of genotypes that represents the genetic space of the breeding 
material well.

\Genetics2level{Coverage of the genetic space}
Animal populations are so large that it does not truly matter, which animals are 
being genotyped once more than 10,000 animals have pedigree records 
\cite{Fragomeni2015,Lourenco2015,Masuda2016}, but in plant breeding programs,
pedigrees are typically considerably smaller.
Single step prediction offers the promise of leveraging the predictive ability
for all genotypes by borrowing the superior performance of an auxiliary
predictor on only a subset of genotypes, thereby reducing predictor costs.
Hence, we were interested in determining which individuals should be 
complemented with information on another predictor.
When using core sets of varying sizes, which were established with the
constraint of maximizing the average genetic distance among their members, we
noticed strong fluctuations in predictive abilities when moving from small core
sets to large core sets.
For traits that were better predicted by transcriptomic than by genomic
information in the core set, predictive abilities typically peaked around
core set size of 0.5/0.6, then dropped sharply at 0.7 and started to increase
again with a size of 0.9.
A principal component analysis showed that, at a core set size of 0.7, 
genotypes assigned to the A3 ancestral population (Fig. \ref{fig:PopStructure})
were largely elements of the core set whereas genotypes with a large share at 
the A1 ancestral population were predominantly excluded from the core set.
At a core set size of 90\% of the original size, the majority of the A1
genotypes were finally covered by the core set as well.
This result suggests a negative bias in the imputation of genotypes from the A1
population and could explain the drop in predictive ability at a core set size
of 0.7, which is in agreement with the reduced predictive abilities observed
when stripping A1 genotypes from their transcriptomic information (Fig.
\ref{fig:InbredResults}B).

>>>>

\section*{Discussion}
\Genetics2level{Advantages of marker effect models}
Single step approaches in hybrid populations have so far been limited to pig
breeding \cite{Xiang2015,Xiang2016,Tusell2016}.
All of these studies used breeding value-based single step BLUP methods
(SS-BLUP) conceived by \citeNP{Legarra2009} and \citeNP{Christensen2010} with
extensions to hybrid populations \cite{Christensen2014,Christensen2015}.
An attactive property of SS-BLUP models is their computational advantage over 
marker-effect single step Bayesian regression models (SSBR) when the number of 
genotypes is smaller than the number of features of the incomplete predictor.
Nevertheless, SS-BLUP suffers from a variety of issues, which we will present
briefly:
First, the combination of the numerator with the genomic relationship matrix
requires commensurability of the two matrices, which necessitates some form of
scaling and weighting of $\mathbf{A}$ and $\mathbf{G}$
\cite{Christensen2012,Christensen2012a}.
Second, the fact that mostly recent individuals have genotypes whereas the 
majority of old individuals has only pedigree records introduces a bias that 
has to be accounted for \cite{Vitezica2011,Legarra2015,Garcia-Baccino2017}.
Third, the addition of new genotypes requires updates of $\mathbf{G}$ that
demand additional measures such as computing the inverse of $\mathbf{G}$ via
recursion
\cite{Misztal2014,Misztal2016,Misztal2016a,Fragomeni2015,Masuda2016,Pocrnic2016}.
Advantages of the SSBR algorithm, conceived by \citeNP{Fernando2014}, include
that i) it does not require commensurability of $\mathbf{A}$ and $\mathbf{G}$, 
ii) it models the bias incurred by the incomplete predictor explicitly as 
$\mathbf{\epsilon}$ and iii) its speed depends largely on the number of 
features.
To our best knowledge, this study presents the first implementation of the SSBR 
algorithm on a hybrid data set.

@@

\Genetics2level{Superiority of single step models}
\subsection{Inbred lines}
When using a single step approach based on complete genomic and incomplete
transcriptomic information, gains in predictive ability for the inbred lines
were at best just slightly higher than when using complete genomic information
alone.
One of the reasons might be that the subset of genotypes that was covered by
transcriptomic data was not a good representation of the full genetic space of
all available genotypes.
Another possibility is that the number of additional phenotypes in the full
set compared to the core set ($\Delta(n) = 62$) was not large enough to provide
considerably more useful information.
The latter was the case for \citeNP{Ashraf2016}, who imputed genotypes for about 
10,000 wheat lines using pedigree information and observed greater accuracy of
the single step method over genomic BLUP for all four evaluated traits.
Predictive abilities reported for the maize diversity panel in this study were
slightly different from those reported by \citeNP{Guo2016}, for three
possible reasons:
1) We reduced the dataset to the 211 tropical/subtropical lines and excluded 
all other genotypes from our analyses whereas \citeNP{Guo2016} used genotypes 
from four different subpopulations, modeled as a fixed effect, and used only a 
subset of 157 genotypes from the set of tropical/subtropical lines.
2) We applied quality checks for the predictor data after generating the subset
of 211 inbred lines, whereas \citeNP{Guo2016} applied the quality checks to 368
genotypes.
3) While \citeNP{Guo2016} used five-fold cross-validation with 500 repetitions,
we employed a leave-one-out cross-validation scheme (LOOCV).
Notwithstanding, the relative differences between predictive abilities for G 
and T, respectively, were similar in both studies.

\subsection{Maize hybrids}
Hybrid breeding --- with noteworthy commercial applications in pig
\cite{Xiang2016,Tusell2016} and maize breeding \cite{TheRoyalSociety2009} --- 
is a particularly challenging application field for prediction tools 
\cite{Kadam2016}.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology \cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.
\citeNP{Westhues2017} have shown that, in hybrid breeding, the probability 
of successfully selecting the best observed genotypes based on the best 
predicted candidates is a strongly convex function of the predictive ability.
Thus, even the minor gains when using transcriptomic and genomic data in a
single step might justify additional investments in RNA-seq at least of a
subset of genotypes that represents the genetic space of the breeding 
material well.
For the hybrid data, the predictive abilities of genomic information
for the whole set of available parent lines rose considerably for several
agronomic traits.
Hence, the inclusion of trancsriptomic data could not further improve upon
this result.
Predictive abilities obtained when using pedigree data could be 
improved considerably when, depending on the agronomic trait, combined with 
either transcriptomic or genomic information.
Evaluations of core sets for the parental inbred lines of the maize hybrids
used throughout this study were outside the scope of this paper, but
results on the inbred lines have shown that homogeneous coverage of the
genetic space by the incomplete predictor can considerably leverage the
predictive ability in single step prediction compared to using only the
complete predictor.

\Genetics2level{Coverage of the genetic space}
Animal populations are so large that it the assembly of a core set of
individuals with records on multiple predictors can effectively be done at 
random once more than 10,000 animals have pedigree records 
\cite{Fragomeni2015,Lourenco2015,Masuda2016}, but in plant breeding programs,
pedigrees are typically considerably smaller.
Single step prediction offers the promise of leveraging the predictive ability
for all genotypes by borrowing the superior performance of an auxiliary
predictor on only a subset of genotypes, thereby reducing predictor costs.
Hence, we were interested in determining which individuals should be 
complemented with information on another predictor.
When using core sets of varying sizes, which were established with the
constraint of maximizing the average genetic distance among their members, we
noticed strong fluctuations in predictive abilities when moving from small core
sets to large core sets.
For traits that were better predicted by transcriptomic than by genomic
information in the core set, predictive abilities typically peaked around
core set size of 0.5/0.6, then dropped sharply at 0.7 and started to increase
again with a size of 0.9.
A principal component analysis showed that, at a core set size of 0.7, 
genotypes assigned to the A3 ancestral population (Fig. \ref{fig:PopStructure})
were largely elements of the core set whereas genotypes with a large share at 
the A1 ancestral population were predominantly excluded from the core set.
At a core set size of 90\% of the original size, the majority of the A1
genotypes were finally covered by the core set as well.
This result suggests a negative bias in the imputation of genotypes from the A1
population and could explain the drop in predictive ability at a core set size
of 0.7, which is in agreement with the reduced predictive abilities observed
when stripping A1 genotypes from their transcriptomic information (Fig.
\ref{fig:InbredResults}B).
